\bf Model Visualization Techniques for a Social Network Model

### 2017-08-18 paper.Rnw
taking a break. will get back to it in the morning.

In this paper, the term \textit{dynamic network} refers to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$. We denote the network observation at timepoint $t_k$ by $x(t_k)$. In the modelling process, we condition on the first observation, $x(t_1)$. The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain (CTMC), which we will denote $X(T)$. This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process,$X(\infty)$, is equivalent to the last observation $x(t_M)$. Nearly all other parts of the process are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$. Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain. Thus, the ``in-between" observations are considered to be ``snapshots" of the network at some point between two steps in the CTMC. The whole process $X(T)$ is a series of single tie changes that happen according to some pre-defined rate function, where one actor at a time is given the opportunity to add or remove one outgoing tie, or to not make any changes. Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize a utility function based on the current and potential future states of the network.

@@

For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$. For this paper, we assume a simple rate \st{"function"}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the model}. \hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter? I updated the language XXX->}%In our model with the simple rate function, 
\st{In our simple model with a rate parameter instead of a rate function,} the rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period from $t_{m}$ to $t_{m+1}$. The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. \st{Let $\tau_i(m)$ be the wait time until actor $i$ makes its next change.For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. The conditional independence assumption is expressed in Equation~\ref{eq:c1}.

\begin{equation} \label{eq:c1}
 \tau_i(m) | x_{i1}(m), \dots, x_{in}(m) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
\end{equation}

 The waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$. The waiting time for the whole network to change, $\tau(m) = \sum_i \tau_i(m) |x_{i1}(m), \dots, x_{in}(m) $ can then be written as

\begin{equation} \label{eq:c2}
 \tau(m) \sim Exp(n\alpha_m)
\end{equation}

} The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs. %In order to achieve the memorylessness property of a Markov process,

@@

\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
Because of the conditional \st{in}dependence assumptions in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, since only one tie from one node is \st{allow to change} at a time. \st{The node $i$, which is the node that is chosen to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alters}. These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point.} So, for the ego node, $i$, in the current network state $x$, the objective function is then written as 
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}

for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$ matrix of covariates. The vector $\boldsymbol{\beta}$ are the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. \hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} \st{XX I added it above the obj function equation. Is it ok there?} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.

The parameters, $\boldsymbol{\beta}$, are attached to various actor-level network statistics, $s_{ik}(x)$. %\hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} 
According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \st{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks. The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software, there are over 80 possible parameters to add to the model. The formulas for the effects are provided in \citet{RSienamanual}. The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation only depend on the structure of the network, like the outdegree and reciprocity parameters mentioned above. \hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} \st{XXX as I'm going through and uncommenting your comments (sorry....) this one has lost meaning a little bit. Could you give me more detail? I'm not sure which section is which at this point. (sorry again...)}
The second set of effects are referred to as the actor-level or covariate effects. These covariate effects also depend on the structure of the network, and are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.

>>>>

In this paper, the term \textit{dynamic network} refers to a network, consisting of a fixed set of $n$ nodes, that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$. We denote the network observation at timepoint $t_k$ by $x(t_k)$. In the modelling process, we condition on the first observation, $x(t_1)$. The SAOM assumes that this longitudinal network of discrete observations is embedded within a continuous time Markov chain (CTMC), which we will denote $X(T)$. This process is almost entirely unobserved: we assume that the beginning of the process, $X(0)$, is equivalent to the first network observation $x(t_1)$, while the end of the process,$X(\infty)$, is equivalent to the last observation $x(t_M)$. Nearly all other parts of the process are unseen, with the exception of $x(t_2), \dots, x(t_{M-1})$. Unlike the first and last observations of the network, these ``in-between" observations do not have direct correspondence with steps in the continuous time Markov chain. Thus, the ``in-between" observations are considered to be ``snapshots" of the network at some point between two steps in the CTMC. The whole process $X(T)$ is a series of single tie changes that happen according to some pre-defined rate function, where one actor at a time is given the opportunity to add or remove one outgoing tie, or to not make any changes. Once an actor is chosen at random according to the rate function, it is ``given" the chance to change a tie, and it tries to maximize its utility function based on the current and near future states of the network. We expand on the model description further in the subsequent sections.

@@

For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\rho(x, \mathbf{z}, \boldsymbol{\alpha})$, where $\boldsymbol{\alpha}$ are the parameters in the function $\rho$, and $x$ is the current network state, with covariates of interest $\mathbf{z}$. For this paper, we assume a simple rate \st{``function"}, $\rho(x, \mathbf{z}, \boldsymbol{\alpha}) = \alpha_m$ that is constant across all actors between observations at time $t_m$ to $t_{m+1}$, \st{thus our rate function is just a rate parameter in the overall model}. In general, SAOMs can incorporate covariate values and network statistics into the model, so that each node will have a different rate of change.  \hh{XXX that means that the function is essentially just a rate parameter, right?} \st{XXX yes XXX}
Other modelling scenarios allow this rate to be more flexible,  e.g.\ a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
\hh{How is the rate parameter different from a rate parameter?} \st{I think you meant how is a rate fn different from a rate parameter? I updated the language XXX->}%In our model with the simple rate function, 
\st{In our simple model with a simple rate parameter instead of a rate function,} the rate parameter dictates how quickly an actor $i$ gets an opportunity to change one of its ties to the other nodes in the network, $x_{ij}$, for $j \in \{1, \dots, n\}$ in the time period from $t_{m}$ to $t_{m+1}$. If $j = i$, no change in the network is made. The model also assumes that the actors $i$ are conditionally independent given their ties, $x_{i1}, \dots, x_{in}$ at the current network state. \st{Let $\tau(i|x,m) $ be the wait time until actor $i$ makes its next change from its current state in the network $x$. Note that $m$ indicates the number of the wave that is conditioned on in the SAOM. For any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. The conditional independence assumption is expressed in Equation~\ref{eq:c1}.

\begin{equation} \label{eq:c1}
 \tau(i|x,m) | x_{i1}(m), \dots, x_{in}(m) \stackrel{\text{iid}}{\sim} Exp(\alpha_m)
\end{equation}

 The waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$. The distribution of waiting time for the whole network to change, $\tau(x|m) = \sum_i \tau_i(m) |x_{i1}(m), \dots, x_{in}(m) $ can then be written as

\begin{equation} \label{eq:c2}
 \tau(x|m) \sim Exp(n\alpha_m)
\end{equation}

} The parameter for the wait time for the whole network $n\alpha_m$ is the rate at which any tie change occurs. The estimation of this parameter is straightforward: a the method of moments is used to estimate the rate with the statistic $$C = \sum_i\sum_j |x_{ij}(t_{m+1}) = x_{ij}(t_m)|$$ which is the total number of changes from observation at time $t_m$ to the observation at time $t_{m+1}$. %In order to achieve the memorylessness property of a Markov process,

@@

\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
Because of the conditional \st{in}dependence assumptions given in Equation~\ref{eq:c1}, we can consider the objective function for each node separately, as only one tie from one node is \st{allowed to change} at a time. \st{The node $i$, which is the node that is chosen to change at the current time point, is called the \textit{ego} node. It has the potential to interact with all other nodes in the network, $j \neq i$. These nodes $j$, are referred to as \textit{alter} nodes, or simply \textit{alters}. These nodes are acted upon by the ego node, and they only act when they become the ego node at a subsequent time point in the CTMC.} For the ego node, $i$, in the current network state $x$, the objective function is written as 
\begin{equation} \label{eq:of1}
f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),
\end{equation}

for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$, the matrix of covariates. The vector $\boldsymbol{\beta}$ contains the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. \hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} \st{XX I added it above the obj function equation. Is it ok there?} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created thatis currently $x_{ij}=0$, or no change will occur.

The parameters, $\boldsymbol{\beta}$, correspond to various actor-level network statistics, $s_{ik}(x)$. \hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} 
According to \citet[p. 371]{snijders01}, there should be at least two parameters included in the model: $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \st{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks. The statistics corresponding to these effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$.} In the \texttt{RSiena} software, there are over 80 possible parameters to add to the model. The formulas for the effects are provided in \citet{RSienamanual}. The parameters, $\beta_k$, in the model can be split up into two groups: first, the structural effects, whose estimation only depend on the structure of the network, like the outdegree and reciprocity parameters mentioned above. \hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} \st{XXX as I'm going through and uncommenting your comments (sorry....) this one has lost meaning a little bit. Could you give me more detail? I'm not sure which section is which at this point. (sorry again...)}
The second set of effects are referred to as the actor-level or covariate effects. These covariate effects also depend on the structure of the network, and are written in terms of the tie variables $x_{ij}$, but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can currently be included in the objective function, see \citet{RSienamanual}.
